Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
debug

@@

@@

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}
```

@@

```{r mice-run}
log_message("Starting MICE imputation process...")
imp_start <- Sys.time()

@@

>>>>

log_message("Variables with missing data:")
print(miss_vars)

# Check that key variables are in the dataset
log_message("\nChecking for key variables in df_mice:")
key_vars_present <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")
for (v in key_vars_present) {
  if (v %in% names(df_mice)) {
    n_miss <- sum(is.na(df_mice[[v]]))
    log_message(paste0(v, ": Present (", n_miss, " missing)"))
  } else {
    log_message(paste0(v, ": NOT PRESENT IN DATA"))
  }
}

@@

log_message("\nDefault methods from mice:")
print(custom_methods)

@@

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}

# DIAGNOSTIC: Verify methods are set correctly
log_message("\nMethods for variables with missing data:")
for (v in miss_vars$Variable) {
  if (v %in% names(custom_methods)) {
    log_message(paste0(v, ": ", custom_methods[v]))
  } else {
    log_message(paste0(v, ": NOT IN METHODS VECTOR"))
  }
}

# DIAGNOSTIC: Check which variables will actually be imputed
vars_to_impute_check <- names(custom_methods)[custom_methods != "" & mice_dry$nmis > 0]
log_message(paste0("\nVariables that SHOULD be imputed (method != '' and has missing): ", 
                   length(vars_to_impute_check)))
print(vars_to_impute_check)
```

@@

```{r mice-run}
log_message("Starting MICE imputation process...")

# DIAGNOSTIC: Final check before imputation
log_message("\nFinal pre-imputation checks:")
log_message(paste0("Number of rows in df_mice: ", nrow(df_mice)))
log_message(paste0("Number of columns in df_mice: ", ncol(df_mice)))
log_message(paste0("Dimensions of predictor matrix: ", nrow(pred_matrix), " x ", ncol(pred_matrix)))
log_message(paste0("Length of methods vector: ", length(custom_methods)))
log_message(paste0("Length of visit sequence: ", length(visit_sequence)))

# Check alignment
if (ncol(df_mice) != length(custom_methods)) {
  log_message("WARNING: Number of columns doesn't match length of methods vector!")
}
if (ncol(df_mice) != nrow(pred_matrix) || ncol(df_mice) != ncol(pred_matrix)) {
  log_message("WARNING: Predictor matrix dimensions don't match data dimensions!")
}

# Show what will be imputed
log_message("\nVariables in visit sequence:")
print(visit_sequence)

# CRITICAL CHECK: Verify outcome_VA_logMAR setup
if ("outcome_VA_logMAR" %in% names(df_mice)) {
  n_miss_outcome <- sum(is.na(df_mice$outcome_VA_logMAR))
  method_outcome <- custom_methods["outcome_VA_logMAR"]
  n_preds_outcome <- sum(pred_matrix["outcome_VA_logMAR", ] > 0)

  log_message("\nCRITICAL: outcome_VA_logMAR status:")
  log_message(paste0("  - Missing values: ", n_miss_outcome))
  log_message(paste0("  - Method assigned: ", method_outcome))
  log_message(paste0("  - Number of predictors: ", n_preds_outcome))
  log_message(paste0("  - In visit sequence: ", "outcome_VA_logMAR" %in% visit_sequence))

  # Check if cohort_id might be blocking
  if ("cohort_id" %in% names(df_mice)) {
    cohort_excluded <- sum(df_mice$cohort_id == "excluded")
    log_message(paste0("  - Records with cohort_id='excluded': ", cohort_excluded))
  }
}

imp_start <- Sys.time()

@@

# Check if imputation actually happened
log_message("\nPost-imputation check - variables that were imputed:")
imputed_vars_actual <- names(mice_results$imp)
log_message(paste0("Number of variables actually imputed: ", length(imputed_vars_actual)))
print(imputed_vars_actual)

# Check specific variables
for (v in c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR")) {
  if (v %in% names(mice_results$imp)) {
    log_message(paste0(v, ": WAS IMPUTED"))
  } else {
    log_message(paste0(v, ": WAS NOT IMPUTED"))
  }
}

# Check for logged events (errors during imputation)
if (!is.null(mice_results$loggedEvents)) {
  log_message("\nWARNING: Logged events during imputation:")
  print(mice_results$loggedEvents)
}
